---
title: "Developing R packages"
author: "Kathryn Doering"
toc: true
format:
  html:
    theme: cerulean
editor: visual
---

## Poll

- Have you ever written documentation in the roxygen format?
- Have you ever run the devtools functions like devtools::document() or devtools::check()?

## Let's dissect an R package!!

Before jumping into writing our own R package, let's check out someone else's.

### Introducing {cowsay}

```{r}
#install.packages("cowsay")
library(cowsay)
say("Welcome to cowsay!")

```

`cowsay::say()` returns an animal (by default, a cat) saying the message you included (yes, that is really it). You can also specify what animal you would like to say the message. `say()`is the only function in this package.

```{r}
say(what = "Let's look at cowsay!", by = "shark")
```

There is also an option to get a fact about cats:

```{r}
say(what = "catfact")
```

There are more examples on the [cowsay website](https://sckott.github.io/cowsay/).

###  Examine the {cowsay} source code

The source code for the cowsay R package [is available on GitHub](https://github.com/sckott/cowsay). We'll take a look at it's anatomy to better understand R packages.

## Design an R package

Our goal will be to build our own R package that uses {cowsay}. Split into groups of ~4 to start designing your R package! Write 2 functions that you want to include in it.

Some things to think about:
- What do you want to name your package?
- Can you write a 2 sentence description of your package?
- What functions would you like to put in the package?
- How could you test your functions?
- Which R packages would you need to rely on (in addition to cowsay)

## Let's build our own R package! 

### Where do we begin?

{devtools} is an essential R package for the R package developer. It provides a diverse set of tools to make it quick and easy to use.

{usethis} provides common r utilities for both r users and developers. In this case, we'll start by running `usethis::create_package()` will create a skeleton R package for you. No need to manually create a bunch of folders!


